---
title: "Dengue raw data cleaner"
format: 
  html:
    toc: true
    output-file: index
  pdf:
    toc: true
    output-file: index
---

# Load libraries
```{r}
.pkgs <- c(
  "tidyverse",
  "janitor",
  "fs",
  "readxl",
  "skimr",
  "stringi",
  "patchwork"
)
xfun::pkg_attach(.pkgs)

theme_set(theme_bw())
```

# Data ingestion
Read raw excel files and quickly clean column names with `janitor::clean_names()`

```{r}
xlsx_raws <- dir_ls("incidence_data_2000_2022", regexp = "xlsx") %>%
  map(
    ~ map(excel_sheets(.x), \(sh) {
      suppressWarnings(read_excel(.x, sheet = sh)) %>% clean_names()
    })
  ) %>%
  flatten() %>%
  set_names(NULL)
```

Our columns of interest are: "sex", "age", "date of admission", "district (of patient's residential address)", "commune (of patient's residential address)", "hospital", "icd", "in-/out-patient"

# EDA and data cleaning
There are 2 excel files, first file contains cases from 2000-2016, a separate sheet for each year. Second find contains cases from 2017-2022, all in one sheet. The 2 files contains 2 different format of columns because they come from 2 different reporting systems. 

## Data from 2000-2016
First we will look at data from 2000-2016

We can look at mismatched columns for all the sheets

```{r}
compare_df_cols(xlsx_raws[-length(xlsx_raws)], return = "mismatch")
```

Mismatched columns `cd_ravien`, `ng_bc`, `ng_tuvong` are insignificant, so we will go ahead with row binding all the sheets together

*`xlsx_raws[-length(xlsx_raws)]` means look at all sheets from first file, ignore last sheet (i.e. second file)*

```{r}
raw_2000_2016 <- xlsx_raws[-length(xlsx_raws)] %>% bind_rows()
```

Skimming to see what's going on in the data
```{r}
skim(raw_2000_2016)
```

The actual columns based on our columns of interest are (with `complete_rate`):
  - `sex` = `gioi` (0.957)
  - `age` = `tuoi` (0.851); `ng_sinh` (0.873); year of birth might be in `hoten` col
  - `date of admission` = `ng_vaovien` (1)
  - `district` = `qh` (1)
  - `commune` = `px` (1)
  - `hospital` = `nguon_du_lieu` (1)
  - `icd` = `m_icd` (0.949)
  - `in-patient` = this is all in-patient data

### Selecting columns of interest

```{r}
s1_2000_2016 <- raw_2000_2016 %>%
  select(hoten, gioi, tuoi, ng_sinh, ng_vaovien, qh, px, nguon_du_lieu, m_icd)
# s1_2000_2016
```

Keeping track of raw number of rows at start to see how much lost during data cleaning
```{r}
start_nrow <- nrow(s1_2000_2016)
start_nrow
```

### Fix `age` if possible
Some year of births (YOBs) are stored in the name (`hoten`) column, let's see how much of the data is like this

```{r}
s1_2000_2016 %>%
  select(hoten, tuoi, ng_sinh) %>%
  filter(is.na(tuoi)) %>%
  rowwise() %>%
  mutate(
    no_age = any(
      varhandle::check.numeric(hoten),
      !is.na(ng_sinh)
    )
  ) %>%
  tabyl(no_age)
```

Less than 1% of the data has YOB in the name column and able to get age from YOB, so it's not worth going deeper.

For now, people with no age will have `NA` as their age

### Rename columns

```{r}
s2_2000_2016 <- s1_2000_2016 %>%
  select(-hoten, -ng_sinh) %>%
  rename(
    sex = gioi,
    age = tuoi,
    date = ng_vaovien,
    district = qh,
    commune = px,
    hospital = nguon_du_lieu,
    icd = m_icd
  ) %>%
  mutate(
    in_out_patient = "in-patient"
  )
# s2_2000_2016
```

### Fix dates data
Date data is in `datetime`, convert to `date` only

```{r}
s3_2000_2016 <- s2_2000_2016 %>%
  mutate(date = convert_to_date(date))
# s3_2000_2016
```


```{r}
skim(s3_2000_2016)
```

We see that there are 7 unique sexes, 37 unique ICDs, negative age. Let's clean all of this

### Fix sexes
Look into `sex` column first

```{r}
s3_2000_2016 %>% tabyl(sex)
```

Impossible to know what "I" and "N" means skip this for now as it's not that important

### Fix ICD

```{r}
s3_2000_2016 %>%
  mutate(year = year(date), icd = tolower(icd)) %>%
  tabyl(icd, year)
```

Needs to consult dengue experts on this, skip for now

### Fix age, again

```{r}
s3_2000_2016 %>% tabyl(age)
```

There are a lot of YOB that are put in as age, let's quickly fix that

```{r}
s3_2000_2016 %>%
  mutate(age = if_else(age > 1000, year(date) - age, age)) %>%
  tabyl(age)
```

Then let's drop negative age and age > 90

```{r}
s4_2000_2016 <- s3_2000_2016 %>%
  mutate(age = if_else(age > 1000, year(date) - age, age)) %>%
  filter(age >= 0, age < 91)

s4_2000_2016 %>% skim()
```

### Wrap up
Probably finished cleaning 2000-2016 data for now

Let's check final number of rows

```{r}
nrow(s4_2000_2016)
start_nrow - nrow(s4_2000_2016)
(start_nrow - nrow(s4_2000_2016)) / start_nrow * 100
```

Lost about 15% of rows

## Data from 2017-2022
Extract the data

```{r}
raw_2017_2022 <- xlsx_raws[length(xlsx_raws)][[1]]
```

Quick skim at the data

```{r}
skim(raw_2017_2022)
```

The actual columns based on our columns of interest are (with `complete_rate`):
  - `sex` = `gioi` (1)
  - `age` = `tuoi` (1.00)
  - `date of admission` = `ngay_nhap_vien` (1)
  - `district` = `quan_huyen_noi_o` (1)
  - `commune` = `phuong_xa_noi_o` (1)
  - `hospital` = `don_vi_bao_cao` (1.00) -- lots of cases out of HCMC -> filter with `tinh_bao_cao` (0.999)
  - `icd` = `phan_do` (0.252)
  - `in-patient` = `tinh_trang_hien_tai` (1.00) -- very complex freetext

### Selecting columns of interest

```{r}
s1_2017_2022 <- raw_2017_2022 %>%
  select(
    gioi,
    tuoi,
    ngay_nhap_vien,
    quan_huyen_noi_o,
    phuong_xa_noi_o,
    don_vi_bao_cao,
    tinh_bao_cao,
    phan_do,
    tinh_trang_hien_tai
  )
s1_2017_2022 %>% skim()
```

Keeping track of raw number of rows at start to see how much lost during data cleaning
```{r}
start_nrow2 <- nrow(s1_2017_2022)
start_nrow2
```

### Filter cases outside of HCMC
First, let's filter out cases that are reported outside of HCMC.

```{r}
s1_2017_2022 %>% tabyl(tinh_bao_cao)
```

Filter based on reporting province
```{r}
s2_2017_2022 <- s1_2017_2022 %>%
  mutate(
    cleaned_tinh_bao_cao = tolower(tinh_bao_cao) %>%
      stri_trans_general(id = "Latin-ASCII") %>%
      str_replace_all("[. -]", "")
  ) %>%
  filter(
    cleaned_tinh_bao_cao %in%
      c(
        "benhviennhidong1",
        "benhviennhidong2",
        "benhvienquan4",
        "bvansinh",
        "hcm",
        "hochiminh",
        "tpcm",
        "tphcm",
        "tphochiminh",
        "tpphcm"
      ) |
      is.na(cleaned_tinh_bao_cao)
  )

s2_2017_2022 %>% tabyl(cleaned_tinh_bao_cao)
```

```{r}
s2_2017_2022 %>% skim()
```

Filter based on reporting hospital

```{r}
s2_2017_2022 %>% tabyl(don_vi_bao_cao)
```

Manual cleaning based on an "eye test" and some quick Googling

```{r}
s2b_2017_2022 <- s2_2017_2022 %>%
  filter(
    !(don_vi_bao_cao %in%
      c(
        "Bệnh viện Lê Lợi - Bà Rịa-V.Tàu",
        "Bệnh viện Quốc tế Becamex",
        "Bệnh viện đa khoa tỉnh Bình Dương",
        "Bệnh viện đa khoa tỉnh Đắk Lắk",
        "PKĐK Phước Lợi - Long An",
        "TTYT Thị xã Trảng Bàng",
        "Trung tâm Y tế  Huyện Vĩnh Hưng",
        "Trung tâm Y tế  Huyện Xuyên Mộc",
        "Trung tâm kiểm soát bệnh tật Long An",
        "Trung tâm y tế tỉnh Trà Vinh"
      ))
  )
s2b_2017_2022 %>% skim()
```

Filter out cases where the reporting hospital is null (only 1 case)

```{r}
s2c_2017_2022 <- s2b_2017_2022 %>% drop_na(don_vi_bao_cao)
s2c_2017_2022 %>% skim()
```

Wrapping this up


```{r}
s3_2017_2022 <- s2c_2017_2022 %>% select(-tinh_bao_cao, -cleaned_tinh_bao_cao)
```

### Rename columns
Let's rename columns before continuing

```{r}
s4_2017_2022 <- s3_2017_2022 %>%
  rename(
    sex = gioi,
    age = tuoi,
    date = ngay_nhap_vien,
    district = quan_huyen_noi_o,
    commune = phuong_xa_noi_o,
    hospital = don_vi_bao_cao,
    icd = phan_do,
    in_out_patient = tinh_trang_hien_tai
  )

s4_2017_2022 %>% skim()
```

### Fix age
We see the same problem with some negative age and YOB put in as age

```{r}
s4_2017_2022 %>% tabyl(age)
```


```{r}
s5_2017_2022 <- s4_2017_2022 %>%
  mutate(age = if_else(age > 2000, year(date) - age, age)) %>%
  filter(age >= 0, age < 91)

s5_2017_2022 %>% tabyl(age)
```

```{r}
s5_2017_2022 %>% skim()
```

### Fix dates
Date data is in `datetime`, convert to `date` only

```{r}
s6_2017_2022 <- s5_2017_2022 %>%
  mutate(date = convert_to_date(date))

s6_2017_2022 %>% skim()
```

### Fix ICD

```{r}
s6_2017_2022 %>%
  mutate(year = year(date), icd = tolower(icd)) %>%
  tabyl(icd, year)
```

Mostly inconsistent letter casing, the classes are rather consistent

```{r}
s7_2017_2022 <- s6_2017_2022 %>%
  mutate(icd = tolower(icd))

s7_2017_2022 %>%
  mutate(year = year(date)) %>%
  tabyl(icd, year)
```

### Fix sexes
```{r}
s7_2017_2022 %>% tabyl(sex)
```

Very easy fix, just remove diacritics, normalise letter casing and recode into english

```{r}
s8_2017_2022 <- s7_2017_2022 %>%
  mutate(sex = stri_trans_general(sex, id = "Latin-ASCII") %>% tolower()) %>%
  mutate(sex = case_when(sex == "nam" ~ "male", sex == "nu" ~ "female"))

s8_2017_2022 %>% tabyl(sex)
```

```{r}
s8_2017_2022 %>% skim()
```

See that there are 25 districts, instead of 24, let's see what's wrong

```{r}
s8_2017_2022 %>% tabyl(district)
```

Just letter casing issue, let's normalise that

### Fix districts

```{r}
s9_2017_2022 <- s8_2017_2022 %>% mutate(district = tolower(district))

s9_2017_2022 %>% skim()
```

### Fix in-/out-patient

Now let's do the hardest part, fixing in-patient and out-patient classification

```{r}
s9_2017_2022 %>% tabyl(in_out_patient)
```

Simplest things to do now are remove diacritics, normalise casing, normaling spacing

```{r}
s9b_2017_2022 <- s9_2017_2022 %>%
  mutate(
    in_out_patient = stri_trans_general(in_out_patient, id = "Latin-ASCII") %>%
      tolower()
  )

s9b_2017_2022 %>%
  tabyl(in_out_patient) %>%
  arrange(desc(n))
```

Lots of random stuff, but major are the in-, out-patient labels and some others (e.g. referred, discharged). The only way to fix this seems to be by hand with regex


```{r}
s10_2017_2022 <- s9b_2017_2022 %>%
  mutate(
    in_out_patient = case_when(
      str_detect(in_out_patient, "(dieu tri)?.*n?g[opa]{2}[ij]{1,2}.*tr?u") ~
        "out-patient",
      str_detect(
        in_out_patient,
        "((di?eu t[tr][ij]{1})?.*no[it]{1}.*tr?[ui])|(n{1,2}hap vien)"
      ) ~
        "in-patient",
      str_detect(
        in_out_patient,
        "((ra|xuat|tron|bo).*vien)|((xin|bo|cho).*(ve|xv))|(ve nha)|(bo kham)"
      ) ~
        "discharged",
      str_detect(in_out_patient, "chuyen.*(benh|bv|vie[nb]|tuyen|khoa)") ~
        "referred",
      .default = "miscellanous"
    )
  )
```

This is as best as I can do...

```{r}
s10_2017_2022 %>%
  tabyl(in_out_patient)
```

```{r}
s10_2017_2022 %>% skim()
```

### Wrap up
Probably finished cleaning 2017-2022 data

Let's check final number of rows

```{r}
nrow(s10_2017_2022)
start_nrow2 - nrow(s10_2017_2022)
(start_nrow2 - nrow(s10_2017_2022)) / start_nrow2 * 100
```

Lost about 2% of rows

# Data joining

Join the 2 data tibles

```{r}
cleaned_incidence_dat <- s4_2000_2016 %>% bind_rows(s10_2017_2022)

cleaned_incidence_dat %>% skim()
```


```{r}
cleaned_incidence_dat %>%
  ggplot() +
  geom_bar(aes(x = date, group = in_out_patient, fill = in_out_patient)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```

## Hospital names

```{r}
cleaned_incidence_dat %>% tabyl(hospital)
```

Normalise hospital names
```{r}
s1_cleaned_incidence_dat <- cleaned_incidence_dat %>%
  mutate(
    hospital = stri_trans_general(hospital, id = "Latin-ASCII") %>% tolower(),
    hospital = gsub("\\s+", " ", hospital) %>% str_replace("bv", "benh vien")
  )

s1_cleaned_incidence_dat %>%
  tabyl(hospital) %>%
  arrange(desc(n))
```

Coding the names of the most busy hospitals
```{r}
s2_cleaned_incidence_dat <- s1_cleaned_incidence_dat %>%
  mutate(
    hospital = case_when(
      str_detect(hospital, "benh vien benh nhiet doi") ~ "HTD",
      str_detect(hospital, "nhi dong 1") ~ "CH1",
      str_detect(hospital, "(nhi dong 2)|nd2") ~ "CH2",
      str_detect(hospital, "nhi dong thanh pho") ~ "CHC",
      str_detect(hospital, "tan phu") ~ "TPH",
      .default = hospital
    )
  )

s2_cleaned_incidence_dat %>%
  tabyl(hospital) %>%
  arrange(desc(n))
```

## Normalise district and commune

Let's normalise district names from the 2 different reporting systems
```{r}
s2_cleaned_incidence_dat %>%
  tabyl(district)
```

```{r}
s3_cleaned_incidence_dat <- s2_cleaned_incidence_dat %>%
  mutate(
    district = stri_trans_general(district, id = "Latin-ASCII") %>% tolower()
  ) %>%
  mutate(district = str_remove(district, "huyen|quan") %>% trimws()) %>%
  mutate(district = trimws(district, which = "left", whitespace = "0"))

s3_cleaned_incidence_dat %>%
  tabyl(district)
```

Now normalise commune names

```{r}
s3_cleaned_incidence_dat %>%
  tabyl(district, commune)
```

```{r}
s3_cleaned_incidence_dat %>%
  filter(between(date, as.Date("2017-01-01"), as.Date("2020-01-01"))) %>%
  tabyl(in_out_patient)
```

## Export to CSV

```{r}
s3_cleaned_incidence_dat %>% write_csv("incidence_full.csv")
```

# Data viz

## Raw time series

```{r eval=FALSE}
raw_ts_plot_dat <- s3_cleaned_incidence_dat %>%
  group_by(across(any_of(c("district", "date")))) %>%
  tally() %>%
  complete(date, fill = list(n = 0)) %>%
  mutate(
    year = year(date),
    isoweek = isoweek(date),
    # isoweek = ifelse(isoweek == 53, 52, isoweek)
  ) %>%
  group_by(year, isoweek) %>%
  summarise(date = min(date), n = sum(n))

raw_ts_plot_dat %>%
  ggplot(aes(x = date, y = n)) +
  geom_step() +
  # facet_wrap(~district, scales = "free_y", ncol = 4) +
  scale_y_continuous("Weekly reported dengue incidence") +
  scale_x_date(
    "",
    date_breaks = "4 years",
    date_labels = "%Y"
  )

raw_ts_plot_dat %>%
  ggplot(aes(x = date, y = n)) +
  # covid lockdown period
  annotate(
    "rect",
    xmin = as.Date("2020/01/01"),
    xmax = as.Date("2022/01/01"),
    ymin = 0,
    ymax = max(raw_ts_plot_dat$n),
    fill = "red",
    alpha = 0.5
  ) +
  geom_vline(
    xintercept = as.Date("2020/01/01"),
    linetype = "dashed",
    alpha = 0.5
  ) +
  annotate(
    "label",
    x = as.Date("2020/01/01"),
    y = 3400,
    label = "COVID-19\nlockdowns",
    color = "#ff4946"
  ) +
  geom_step() +
  # circ 54
  annotate(
    "segment",
    x = as.Date("2018/01/01"),
    y = 0,
    yend = 2800,
    linetype = 2,
    color = "blue"
  ) +
  annotate(
    "label",
    x = as.Date("2018/01/01"),
    y = 2800,
    label = "Circular\n54/2017/TT-BYT\nin effect",
    color = "blue"
  ) +
  scale_y_continuous("") +
  scale_x_date(
    "",
    date_breaks = "2 years",
    date_minor_breaks = "1 year",
    date_labels = "%Y"
  ) +
  ggtitle("Weekly reported dengue incidence in HCMC")

ggsave("plots/raw_ts_plot.jpg", width = 10)
```

## Incidence by case status

```{r}
case_status_plot_dat <- s3_cleaned_incidence_dat %>%
  group_by(date, in_out_patient) %>%
  tally() %>%
  ungroup() %>%
  complete(date, in_out_patient, fill = list(n = 0)) %>%
  mutate(
    year = year(date),
    isoweek = isoweek(date),
    # isoweek = ifelse(isoweek == 53, 52, isoweek)
  ) %>%
  group_by(year, isoweek, in_out_patient) %>%
  summarise(date = min(date), n = sum(n))

p1 <- case_status_plot_dat %>%
  ggplot(aes(x = date, y = n, fill = in_out_patient)) +
  geom_area() +
  ggtitle("Weekly dengue incidence by case status")

p2 <- case_status_plot_dat %>%
  group_by(date) %>%
  mutate(n_pct = n / sum(n)) %>%
  ggplot(aes(x = date, y = n_pct, fill = in_out_patient)) +
  geom_area() +
  ggtitle("Percentage of weekly dengue incidence by case status")

(p1 / p2) +
  plot_layout(guides = "collect", axes = "collect") &
  geom_vline(
    xintercept = as.Date("2020/01/01"),
    linetype = "dashed",
    alpha = 0.5
  ) &
  scale_fill_brewer(
    "Case status",
    labels = str_to_title,
    palette = "Set1"
  ) &
  scale_x_date(
    "",
    date_breaks = "2 years",
    date_minor_breaks = "1 year",
    date_labels = "%Y"
  ) &
  scale_y_continuous("")


ggsave("plots/case_status_plot.jpg", width = 10)
```

## Total number of cases per hospital

```{r}
hospital_order <- s3_cleaned_incidence_dat %>%
  mutate(week = lubridate::floor_date(date, "week")) %>%
  group_by(hospital) %>%
  tally(name = "total_n") %>%
  arrange(desc(total_n)) %>%
  pull(hospital)


s3_cleaned_incidence_dat %>%
  filter(in_out_patient %in% c("in-patient", "out-patient")) %>%
  mutate(
    week = lubridate::floor_date(date, "week"),
    # hospitals take make up less the 5% of total cases will be put in "Other"
    hospital = factor(hospital, levels = hospital_order) %>% fct_lump_prop(0.05)
  ) %>%
  group_by(week, in_out_patient, hospital) %>%
  tally() %>%
  ggplot(aes(x = week, y = n, color = in_out_patient)) +
  geom_line() +
  geom_point(size = 0.5) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    minor_breaks = NULL
  ) +
  #   facet_wrap(~hospital, ncol = 1)
  facet_wrap(~hospital, ncol = 1, scales = "free_y")
#   theme(legend.position = "none")
```

We can see some weird period where there are so little in-patient data from big hospitals like in 2017.

Let's zoom into that

```{r}
s3_cleaned_incidence_dat %>%
  filter(year(date) == 2017, in_out_patient == "in-patient") %>%
  tabyl(hospital) %>%
  arrange(desc(n))

s3_cleaned_incidence_dat %>%
  filter(year(date) == 2017, in_out_patient == "in-patient") %>%
  mutate(
    week = lubridate::floor_date(date, "week"),
    hospital = factor(hospital, levels = hospital_order) %>% fct_lump_prop(0.05)
  ) %>%
  group_by(week, in_out_patient, hospital) %>%
  tally() %>%
  ggplot(aes(x = week, y = n, color = in_out_patient)) +
  geom_line() +
  geom_point(size = 0.5) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    minor_breaks = NULL
  ) +
  facet_wrap(~hospital, ncol = 1) +
  #   facet_wrap(~hospital, ncol = 1, scales = "free_y")
  theme(legend.position = "none")
```

## Data availability map

```{r}
s3_cleaned_incidence_dat %>%
  mutate(week = floor_date(date, "week")) %>%
  group_by(week, in_out_patient, hospital) %>%
  tally() %>%
  ungroup() %>%
  complete(week, in_out_patient, hospital) %>%
  filter(in_out_patient %in% c("in-patient", "out-patient")) %>%
  group_by(week, in_out_patient, hospital) %>%
  mutate(total_n = sum(n, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(hospital = fct_lump_prop(hospital, prop = 0.01, w = total_n)) %>%
  ggplot(aes(x = week, y = fct_reorder(hospital, total_n), fill = n)) +
  geom_raster() +
  facet_wrap(~in_out_patient, ncol = 1) +
  theme(legend.position = "none") +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    minor_breaks = NULL
  ) +
  scale_fill_viridis_c(na.value = "transparent")

```