---
title: "Dengue raw data cleaner"
format: 
  html:
    toc: true
    output-file: index
  pdf:
    toc: true
    output-file: index
editor: 
  markdown: 
    wrap: 72
---

# Load libraries

```{r}
.pkgs <- c(
  "tidyverse",
  "janitor",
  "fs",
  "readxl",
  "skimr",
  "stringi",
  "patchwork"
)
xfun::pkg_attach(.pkgs)

theme_set(theme_bw())
```

# Data ingestion

Read raw excel files and quickly clean column names with
`janitor::clean_names()`

```{r}
xlsx_raws <- list.files(
  "incidence_data_geocoded",
  pattern = "xlsx",
  full.names = TRUE
) %>%
  map(
    ~ map(excel_sheets(.x), \(sh) {
      suppressWarnings(read_excel(.x, sheet = sh)) %>% clean_names()
    })
  ) %>%
  set_names(NULL)
```

Our columns of interest are: "sex", "age", "date of admission",
"district (of patient's residential address)", "commune (of patient's
residential address)", "hospital", "icd", "in-/out-patient"

# EDA and data cleaning

There are 3 excel files

-   First file contains cases from 2000-2016, a separate sheet for each
    year.

-   Second file contains cases from 2017-2025, a separate sheet for each
    year

-   Third file contains cases from August 2025 (after major change in
    admin level).

## Data from 2000-2016

We can look at mismatched columns for all the sheets

```{r}
compare_df_cols(xlsx_raws[[1]], return = "mismatch")
```

Mismatched columns `cd_ravien`, `ng_bc`, `ng_tuvong` are insignificant,
so we will go ahead with row binding all the sheets together

*`xlsx_raws[-length(xlsx_raws)]` means look at all sheets from first
file, ignore last sheet (i.e. second file)*

```{r}
raw_2000_2016 <- xlsx_raws[[1]] %>% bind_rows()
```

Skimming to see what's going on in the data

```{r}
skim(raw_2000_2016)
```

The actual columns based on our columns of interest are (with
`complete_rate`):

-   `sex` = `gioi` (0.957)

-   `age` = `tuoi` (0.851); `ng_sinh` (0.783); year of birth might be in
    `hoten` col

-   `date of admission` = `ng_vaovien` (1)

-   `district` = `qh` (1)

-   `commune` = `px` (1)

-   `new_commune` = `px_moi` (0.904)

-   `hospital` = `nguon_du_lieu` (1)

-   `icd` = `m_icd` (0.949)

-   `in-patient` = this is all in-patient data

### Selecting columns of interest

```{r}
s1_2000_2016 <- raw_2000_2016 %>%
  select(
    hoten,
    gioi,
    tuoi,
    ng_sinh,
    ng_vaovien,
    qh,
    px,
    px_moi,
    nguon_du_lieu,
    m_icd
  )
# s1_2000_2016
```

Keeping track of raw number of rows at start to see how much lost during
data cleaning

```{r}
start_nrow <- nrow(s1_2000_2016)
start_nrow
```

### Fix `age` if possible

Some year of births (YOBs) are stored in the name (`hoten`) column,
let's see how much of the data is like this

```{r}
s1_2000_2016 %>%
  select(hoten, tuoi, ng_sinh) %>%
  filter(is.na(tuoi)) %>%
  rowwise() %>%
  mutate(
    no_age = any(
      varhandle::check.numeric(hoten),
      !is.na(ng_sinh)
    )
  ) %>%
  tabyl(no_age)
```

Less than 1% of the data has YOB in the name column and able to get age
from YOB, so it's not worth going deeper.

For now, people with no age will have `NA` as their age

### Rename columns

```{r}
s2_2000_2016 <- s1_2000_2016 %>%
  select(-hoten, -ng_sinh) %>%
  rename(
    sex = gioi,
    age = tuoi,
    date = ng_vaovien,
    district = qh,
    commune = px,
    new_commune = px_moi,
    hospital = nguon_du_lieu,
    icd = m_icd
  ) %>%
  mutate(
    in_out_patient = "in-patient"
  )
# s2_2000_2016
```

### Fix dates data

Date data is in `datetime`, convert to `date` only

```{r}
s3_2000_2016 <- s2_2000_2016 %>%
  mutate(date = convert_to_date(date))
# s3_2000_2016
```

```{r}
skim(s3_2000_2016)
```

We see that there are 7 unique sexes, 37 unique ICDs, negative age.
Let's clean all of this

### Fix sexes

Look into `sex` column first

```{r}
s3_2000_2016 %>% tabyl(sex)
```

Impossible to know what "I" and "N", treat these as NA instead

Relabel `sex` column

```{r}
s3_2000_2016 <- s3_2000_2016 |>
  mutate(
    sex = tolower(sex),
    sex = case_when(
      sex == "g" ~ "f",
      sex == "t" ~ "m",
      .default = NA
    )
  )
```

### Fix ICD

```{r}
s3_2000_2016 %>%
  mutate(year = year(date), icd = tolower(icd)) %>%
  tabyl(icd, year)
```

Needs to consult dengue experts on this, skip for now

### Fix age, again

```{r}
s3_2000_2016 %>% tabyl(age)
```

There are a lot of YOB that are put in as age, let's quickly fix that

```{r}
s3_2000_2016 %>%
  mutate(age = if_else(age > 1000, year(date) - age, age)) %>%
  tabyl(age)
```

Then let's drop negative age and age \> 90

```{r}
s4_2000_2016 <- s3_2000_2016 %>%
  mutate(age = if_else(age > 1000, year(date) - age, age)) %>%
  filter(age >= 0, age < 91)

s4_2000_2016 %>% skim()
```

### Wrap up

Probably finished cleaning 2000-2016 data for now

Let's check final number of rows

```{r}
nrow(s4_2000_2016)
start_nrow - nrow(s4_2000_2016)
(start_nrow - nrow(s4_2000_2016)) / start_nrow * 100
```

Lost about 15% of rows

## Data from 2017-2025

Extract the data

```{r}
raw_2017_2025 <- xlsx_raws[[2]] |> bind_rows()
```

Quick skim at the data

```{r}
skim(raw_2017_2025)
```

The actual columns based on our columns of interest are (with
`complete_rate`):

-   `age` = `tuoi` (0.999)

-   `date of admission` = `ngay_nhap_vien_kham_benh` (1)

-   `district` = `quan_huyen_cu` (1)

-   `commune` = `phuong_xa_cu` (1)

-   `new_commune` = `px_moi` (0.975)

-   `icd` = `phan_do_benh` (0.252)

-   `in-patient` = `hinh_thuc_dieu_tri` (1.00)

### Selecting columns of interest

```{r}
s1_2017_2025 <- raw_2017_2025 %>%
  select(
    tuoi,
    ngay_nhap_vien_kham_benh,
    quan_huyen_cu,
    phuong_xa_cu,
    px_moi,
    phan_do_benh,
    hinh_thuc_dieu_tri
  )
s1_2017_2025 %>% skim()
```

Keeping track of raw number of rows at start to see how much lost during
data cleaning

```{r}
start_nrow2 <- nrow(s1_2017_2025)
start_nrow2
```

### Rename columns

Let's rename columns before continuing

```{r}
s2_2017_2025 <- s1_2017_2025 %>%
  rename(
    age = tuoi,
    date = ngay_nhap_vien_kham_benh,
    district = quan_huyen_cu,
    commune = phuong_xa_cu,
    new_commune = px_moi,
    icd = phan_do_benh,
    in_out_patient = hinh_thuc_dieu_tri
  )

s2_2017_2025 %>% skim()
```

### Fix age

We see the same problem with some negative age and YOB put in as age

```{r}
s2_2017_2025 %>% tabyl(age)
```

```{r}
s3_2017_2025 <- s2_2017_2025 %>%
  mutate(age = if_else(age > 2000, year(date) - age, age)) %>%
  filter(age >= 0, age < 91)

s3_2017_2025 %>% tabyl(age)
```

```{r}
s3_2017_2025 %>% skim()
```

### Fix dates

Date data is in `datetime`, convert to `date` only

```{r}
s4_2017_2025 <- s3_2017_2025 %>%
  mutate(date = convert_to_date(date))

s4_2017_2025 %>% skim()
```

### Fix ICD

```{r}
s4_2017_2025 %>%
  mutate(year = year(date), icd = tolower(icd)) %>%
  tabyl(icd, year)
```

Mostly inconsistent letter casing, the classes are rather consistent

```{r}
s5_2017_2025 <- s4_2017_2025 %>%
  mutate(icd = tolower(icd))

s5_2017_2025 %>%
  mutate(year = year(date)) %>%
  tabyl(icd, year)
```

### Fix districts

```{r}
s6_2017_2025 <- s5_2017_2025 %>% mutate(district = tolower(district))

s6_2017_2025 %>% skim()
```

### Fix in-/out-patient

Now let's fix in-patient and out-patient classification

```{r}
s6_2017_2025 %>% tabyl(in_out_patient)
```

Simplest things to do now are remove diacritics, normalise casing,
normaling spacing

```{r}
s7_2017_2025 <- s6_2017_2025 %>%
  mutate(
    in_out_patient = stri_trans_general(in_out_patient, id = "Latin-ASCII") %>%
      tolower()
  )

s7_2017_2025 %>%
  tabyl(in_out_patient) %>%
  arrange(desc(n))
```

Relabel `in_out_patient` status

```{r}
s8_2017_2025 <- s7_2017_2025 %>%
  mutate(
    in_out_patient = case_when(
      in_out_patient == "noi tru" ~ "in-patient",
      in_out_patient == "ngoai tru" ~ "out-patient",
      .default = "other"
    )
  )
```

```{r}
s8_2017_2025 %>%
  tabyl(in_out_patient)
```

```{r}
s8_2017_2025 %>% skim()
```

### Wrap up

Probably finished cleaning 2017-2025 data

Let's check final number of rows

```{r}
nrow(s8_2017_2025)
start_nrow2 - nrow(s8_2017_2025)
(start_nrow2 - nrow(s8_2017_2025)) / start_nrow2 * 100
```

Lost less than 1% of rows

## Data after the 2025 reform

Extract the data

```{r}
raw_aug2025 <- xlsx_raws[[3]] |> bind_rows()
```

Quick skim at the data

```{r}
skim(raw_aug2025)
```

The actual columns based on our columns of interest are (with
`complete_rate`):

-   `age` = `tuoi` (1)

-   `date of admission` = `ngay_nhap_vien_kham_benh` (1)

-   `district` = `quan_huyen_cu` (0.804)

-   `commune` = `phuong_xa_cu` (0.804)

-   `new_commune` = `phuong_xa_moi` (1)

-   `icd` = `phan_do_benh` (1)

-   `in-patient` = `hinh_thuc_dieu_tri` (1.00)

### Selecting columns of interest

```{r}
s1_aug2025 <- raw_aug2025 %>%
  select(
    tuoi,
    ngay_nhap_vien_kham_benh,
    quan_huyen_cu,
    phuong_xa_cu,
    phuong_xa_moi,
    phan_do_benh,
    hinh_thuc_dieu_tri
  )
s1_aug2025 %>% skim()
```

Keeping track of raw number of rows at start to see how much lost during
data cleaning

```{r}
start_nrow3 <- nrow(s1_aug2025)
start_nrow3
```

### Rename columns

Let's rename columns before continuing

```{r}
s2_aug2025 <- s1_aug2025 %>%
  rename(
    age = tuoi,
    date = ngay_nhap_vien_kham_benh,
    district = quan_huyen_cu,
    commune = phuong_xa_cu,
    new_commune = phuong_xa_moi,
    icd = phan_do_benh,
    in_out_patient = hinh_thuc_dieu_tri
  )

s2_aug2025 %>% skim()
```

### Fix age

Age looks quite good in this case, no YoB in data and no negative value

However, filter out cases where age \> 91 to be consistent

```{r}
s2_aug2025 %>% tabyl(age)
```

```{r}
s3_aug2025 <- s2_aug2025 %>%
  filter(age >= 0, age < 91)

s3_aug2025 %>% tabyl(age)
```

```{r}
s3_aug2025 %>% skim()
```

### Fix dates

Date data is in `datetime`, convert to `date` only

```{r}
s4_aug2025 <- s3_aug2025 %>%
  mutate(date = convert_to_date(date))

s4_aug2025 %>% skim()
```

### Fix ICD

```{r}
s4_aug2025 %>%
  mutate(year = year(date), icd = tolower(icd)) %>%
  tabyl(icd, year)
```

Mostly inconsistent letter casing, the classes are rather consistent

```{r}
s5_aug2025 <- s4_aug2025 %>%
  mutate(icd = tolower(icd))

s5_aug2025 %>%
  mutate(year = year(date)) %>%
  tabyl(icd, year)
```

### Fix districts

```{r}
s6_aug2025 <- s5_aug2025 %>% mutate(district = tolower(district))

s6_aug2025 %>% skim()
```

### Fix in-/out-patient

Now let's fix in-patient and out-patient classification

```{r}
s6_aug2025 %>% tabyl(in_out_patient)
```

```{r}
s7_aug2025 <- s6_aug2025 %>%
  mutate(
    in_out_patient = stri_trans_general(in_out_patient, id = "Latin-ASCII") %>%
      tolower()
  )

s7_aug2025 %>%
  tabyl(in_out_patient) %>%
  arrange(desc(n))
```

Relabel `in_out_patient` status

```{r}
s8_aug2025 <- s7_aug2025 %>%
  mutate(
    in_out_patient = case_when(
      in_out_patient == "noi tru" ~ "in-patient",
      in_out_patient == "ngoai tru" ~ "out-patient",
      .default = "other"
    )
  )
```

```{r}
s8_aug2025 %>%
  tabyl(in_out_patient)
```

```{r}
s8_aug2025 %>% skim()
```

### Wrap up

Check final number of rows

```{r}
nrow(s8_aug2025)
start_nrow3 - nrow(s8_aug2025)
(start_nrow3 - nrow(s8_aug2025)) / start_nrow3 * 100
```

# Data joining

For now, we only need incidence from `2000-2020` with the label for the
corresponding new commune

```{r}
cleaned_incidence_dat <- s4_2000_2016 |>
  bind_rows(s8_2017_2025) |>
  filter(date < as.Date("2020-01-01"))
cleaned_incidence_dat %>% skim()
```

```{r}
cleaned_incidence_dat %>%
  ggplot() +
  geom_bar(aes(x = date, group = in_out_patient, fill = in_out_patient)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```

## Normalise district and commune

Let's normalise district names from the 2 different reporting systems

```{r}
cleaned_incidence_dat %>%
  tabyl(district)
```

```{r}
s2_cleaned_incidence_dat <- cleaned_incidence_dat %>%
  mutate(
    district = stri_trans_general(district, id = "Latin-ASCII") %>% tolower()
  ) %>%
  mutate(district = str_remove(district, "huyen|quan") %>% trimws()) %>%
  mutate(district = trimws(district, which = "left", whitespace = "0"))

s2_cleaned_incidence_dat %>%
  tabyl(district) |>
  dim()
```

Now normalise commune names

```{r}
s2_cleaned_incidence_dat %>%
  tabyl(district, commune)
```

```{r}
s3_cleaned_incidence_dat <- s2_cleaned_incidence_dat |>
  mutate(
    commune = stri_trans_general(commune, id = "Latin-ASCII") %>% tolower(),
    commune = str_remove(commune, "phuong|xa|thi tran") %>% trimws(),
    commune = trimws(commune, which = "left", whitespace = "0"),
    commune = case_when(
      commune == "khong ro" ~ NA,
      commune == "" ~ NA,
      .default = commune
    )
  )
```

```{r}
s4_cleaned_incidence_dat <- s3_cleaned_incidence_dat |>
  select(sex, age, date, district, commune, new_commune, in_out_patient) |>
  rename(
    date_admitted = date,
    status = in_out_patient
  )
```

## Export to CSV

```{r}
s4_cleaned_incidence_dat |>
  arrange(date_admitted) |>
  write_csv("out/incidence_clean_2000_2020.csv")
```

# Data viz

## Raw time series

```{r eval=FALSE}
raw_ts_plot_dat <- s4_cleaned_incidence_dat %>%
  group_by(across(any_of(c("district", "date_admitted")))) %>%
  tally() %>%
  complete(date_admitted, fill = list(n = 0)) %>%
  mutate(
    year = year(date_admitted),
    isoweek = isoweek(date_admitted),
    # isoweek = ifelse(isoweek == 53, 52, isoweek)
  ) %>%
  group_by(year, isoweek) %>%
  summarise(date = min(date_admitted), n = sum(n))

raw_ts_plot_dat %>%
  ggplot(aes(x = date, y = n)) +
  geom_step() +
  # facet_wrap(~district, scales = "free_y", ncol = 4) +
  scale_y_continuous("Weekly reported dengue incidence") +
  scale_x_date(
    "",
    date_breaks = "4 years",
    date_labels = "%Y"
  )

raw_ts_plot_dat %>%
  ggplot(aes(x = date, y = n)) +
  # covid lockdown period
  annotate(
    "rect",
    xmin = as.Date("2020/01/01"),
    xmax = as.Date("2022/01/01"),
    ymin = 0,
    ymax = max(raw_ts_plot_dat$n),
    fill = "red",
    alpha = 0.5
  ) +
  geom_vline(
    xintercept = as.Date("2020/01/01"),
    linetype = "dashed",
    alpha = 0.5
  ) +
  annotate(
    "label",
    x = as.Date("2020/01/01"),
    y = 3400,
    label = "COVID-19\nlockdowns",
    color = "#ff4946"
  ) +
  geom_step() +
  # circ 54
  annotate(
    "segment",
    x = as.Date("2018/01/01"),
    y = 0,
    yend = 2800,
    linetype = 2,
    color = "blue"
  ) +
  annotate(
    "label",
    x = as.Date("2018/01/01"),
    y = 2800,
    label = "Circular\n54/2017/TT-BYT\nin effect",
    color = "blue"
  ) +
  scale_y_continuous("") +
  scale_x_date(
    "",
    date_breaks = "2 years",
    date_minor_breaks = "1 year",
    date_labels = "%Y"
  ) +
  ggtitle("Weekly reported dengue incidence in HCMC")

ggsave("plots/raw_ts_plot.jpg", width = 10)
```

## Incidence by case status

```{r}
case_status_plot_dat <- s4_cleaned_incidence_dat %>%
  group_by(date_admitted, status) %>%
  tally() %>%
  ungroup() %>%
  complete(date_admitted, status, fill = list(n = 0)) %>%
  mutate(
    year = year(date_admitted),
    isoweek = isoweek(date_admitted),
    # isoweek = ifelse(isoweek == 53, 52, isoweek)
  ) %>%
  group_by(year, isoweek, status) %>%
  summarise(date = min(date_admitted), n = sum(n))

p1 <- case_status_plot_dat %>%
  ggplot(aes(x = date, y = n, fill = status)) +
  geom_area() +
  ggtitle("Weekly dengue incidence by case status")

p2 <- case_status_plot_dat %>%
  group_by(date) %>%
  mutate(n_pct = n / sum(n)) %>%
  ggplot(aes(x = date, y = n_pct, fill = status)) +
  geom_area() +
  ggtitle("Percentage of weekly dengue incidence by case status")

(p1 / p2) +
  plot_layout(guides = "collect", axes = "collect") &
  geom_vline(
    xintercept = as.Date("2020/01/01"),
    linetype = "dashed",
    alpha = 0.5
  ) &
  scale_fill_brewer(
    "Case status",
    labels = str_to_title,
    palette = "Set1"
  ) &
  scale_x_date(
    "",
    date_breaks = "2 years",
    date_minor_breaks = "1 year",
    date_labels = "%Y"
  ) &
  scale_y_continuous("")


ggsave("plots/case_status_plot.jpg", width = 10)
```

## Data availability map

```{r}
s4_cleaned_incidence_dat %>%
  mutate(week = floor_date(date_admitted, "week")) %>%
  group_by(week, status, district) %>%
  tally() %>%
  ungroup() %>%
  complete(week, status, district) %>%
  filter(status %in% c("in-patient", "out-patient")) %>%
  group_by(week, status, district) %>%
  mutate(total_n = sum(n, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(hospital = fct_lump_prop(district, prop = 0.01, w = total_n)) %>%
  ggplot(aes(x = week, y = fct_reorder(district, total_n), fill = n)) +
  geom_raster() +
  facet_wrap(~status, ncol = 1) +
  theme(legend.position = "none") +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y",
    minor_breaks = NULL
  ) +
  scale_fill_viridis_c(na.value = "transparent")
```
